---
title: "Movement-State Segmentation with BALM"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Movement-State Segmentation with BALM}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

## Overview

`balm_segmentation()` classifies each tracking fix into a movement state
using **BALM** (Behaviorally-Anchored Local Moran), a movement-adapted
variant of the Local Moran scatterplot. It uses only point locations and a
movement-rate mark -- never direction or the raw temporal sequence -- so it
is invariant to fix ordering and robust to sampling-rate heterogeneity.

Each fix is labelled by the signs of its deviation from a behavioral
reference `c` (the migratory-flight onset) and its spatial lag:

- **HH** -- fast fix among fast neighbours -> active migration
- **LL** -- slow fix among slow neighbours -> stationary site
- **HL** -- fast fix among slow neighbours -> local movement within a site (commuting / foraging burst)
- **LH** -- slow fix among fast neighbours -> brief touch-down in the migratory corridor (transient stopover)

Two choices distinguish BALM from the textbook Local Moran: the high/low
reference is a **data-driven behavioral threshold** (the flight-onset
speed) rather than the arithmetic mean, and classification is
**deterministic** (no permutation significance) -- spatial support is
arbitrated downstream by density clustering (`dbscan_habitats()`).

## Setup

```{r}
library(movepp)
library(sf)
library(ggplot2)
data(godwit_demo)
```

## Movement rate

```{r}
godwit <- compute_step_speed(godwit_demo,
                             time_col       = "time",
                             individual_col = "individual",
                             direction      = "centered")
```

## Segmentation

```{r}
seg <- balm_segmentation(godwit,
                         variable_col   = "step_speed",
                         individual_col = "individual",
                         verbose        = FALSE)
table(seg$cluster_type, useNA = "ifany")
attr(seg, "balm_reference")   # flight-onset reference c (km/h)
```

## Where the high/low reference comes from

The reference `c` is the **flight-onset speed** -- detected as the left
foot of the right-most peak of the (log) step-speed density, i.e. the
boundary between site-level movement and migratory flight.

```{r ref-plot, fig.width=7, fig.height=4}
ref <- attr(seg, "balm_reference")[[1]]
spd <- godwit$step_speed[is.finite(godwit$step_speed) & godwit$step_speed > 0]

ggplot(data.frame(spd = spd), aes(spd)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "grey85", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = ref, colour = "#D7191C", linewidth = 1) +
  scale_x_log10() +
  labs(x = "step speed (km/h, log scale)", y = "density",
       title = sprintf("Behavioral reference: flight onset c = %.1f km/h", ref),
       subtitle = "Red line separates site-level movement from migratory flight") +
  theme_minimal(base_size = 12)
```

## The BALM Moran scatterplot

Each fix is placed by its own deviation `x - c` (x-axis) against the mean
deviation of its spatial neighbours (y-axis); the four quadrants are the
four movement states.

```{r scatter, fig.width=6.5, fig.height=5}
cluster_pal <- c(HH = "#F0B8B1", LL = "#99D0E8", HL = "#E01B1B", LH = "#1B53E0")

ggplot(sf::st_drop_geometry(seg),
       aes(balm_deviation, balm_lag, colour = cluster_type)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(alpha = 0.5, size = 0.7) +
  scale_colour_manual(values = cluster_pal, name = "state", na.translate = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "deviation from flight onset  (x - c)",
       y = "spatial lag  (mean neighbour deviation)",
       title = "BALM Moran scatterplot") +
  theme_minimal(base_size = 12)
```

## Spatial result

```{r map, fig.width=6.5, fig.height=7}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
bb    <- sf::st_bbox(seg)
padx  <- as.numeric(bb["xmax"] - bb["xmin"]) * 0.10
pady  <- as.numeric(bb["ymax"] - bb["ymin"]) * 0.10

ggplot() +
  geom_sf(data = world, fill = "grey96", colour = "grey80", linewidth = 0.2) +
  geom_sf(data = seg, aes(colour = cluster_type), size = 0.7, alpha = 0.8) +
  scale_colour_manual(values = cluster_pal, name = "state", na.translate = FALSE) +
  coord_sf(xlim = c(bb["xmin"] - padx, bb["xmax"] + padx),
           ylim = c(bb["ymin"] - pady, bb["ymax"] + pady), expand = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  labs(title = "Movement states across geographic space") +
  theme_minimal(base_size = 12)
```

## Stationary subset for downstream analysis

The `LL` (stationary cores) and `LH` (transient stopovers) fixes are the
input to spatial habitat delineation:

```{r}
stationary <- seg[!is.na(seg$cluster_type) &
                  seg$cluster_type %in% c("LL", "LH"), ]
nrow(stationary)
```

## Discussion

BALM asks "is this fix part of a spatial cluster of high (or low) movement
rate", using spatial structure alone. The behavioral signature of
migration -- fast, transient passage -- manifests spatially as high-rate
fixes in sparse corridors (`HH`), whereas fast movements confined to a site
appear as `HL` outliers; no explicit measure of direction is required.

## See also

- `vignette("v01-data-overview")` -- full workflow
- `vignette("v03-dbscan-habitats")` -- downstream clustering of LL/LH points

## Session info

```{r session-info}
sessionInfo()
```
