---
title: "Inter-Annual Site Fidelity (ISFI)"
format:
  html:
    embed-resources: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Inter-Annual Site Fidelity (ISFI)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

# Overview

`compute_isfi()` quantifies how often an individual returns to the
same temporary habitat (or nest) across years:

$$
\text{ISFI} = \frac{\sum_i n_{i,\text{years}} - n}{(c - 1) \cdot n}
$$

where `n` is unique clusters, `c` is unique years, and
`n_{i,years}` is the number of years cluster `i` was visited.
ISFI ranges from 0 (no fidelity) to 1 (every cluster revisited every
year).

A single overall ISFI averaged across an entire annual cycle is
biologically uninformative: a bird may be highly faithful to its
wintering grounds but choose different stopover sites each year.
The recommended workflow is therefore to **stratify ISFI by life-
history phase**, exposing where in the cycle fidelity is strongest.

# Setup

```{r}
library(movepp)
library(sf)
library(ggplot2)
```

# Build phase-labelled habitats

We feed `godwit_demo` through the full macro-meso pipeline
(step-speed → BALM segmentation → DBSCAN → MPI → density-valley
phase classification) to obtain a set of temporary habitats each
labelled with a life-history phase:

```{r}
data("godwit_demo")

godwit <- compute_step_speed(godwit_demo,
                              time_col       = "time",
                              individual_col = "individual",
                              direction      = "centered")

seg <- balm_segmentation(godwit,
                          variable_col    = "step_speed",
                          individual_col  = "individual",
                          verbose         = FALSE)

stationary   <- seg[!is.na(seg$cluster_type) & seg$cluster_type %in% c("LL", "LH"), ]
p    <- detect_habitat_params(stationary, track = seg,
                              individual_col = "individual", time_col = "time")
hab  <- dbscan_habitats(stationary,
                        individual_col = "individual",
                        eps = p$eps, minPts = p$minPts)
mpi  <- compute_mpi(hab, individual_col = "individual",
                    time_col = "time", verbose = FALSE)
phs  <- classify_phases(mpi, n_phases = 3, individual_col = "individual")
```

`classify_phases()` returns phase labels ordered by ascending
mean MPI. We remap these to biologically meaningful names:

```{r}
ph_df    <- sf::st_drop_geometry(phs)
ph_order <- names(sort(tapply(ph_df$mpi, ph_df$phase, mean, na.rm = TRUE)))
phase_map <- setNames(c("wintering", "stopover", "breeding"), ph_order)
phs$phase_named <- factor(phase_map[as.character(phs$phase)],
                          levels = c("wintering", "stopover", "breeding"))

table(phs$phase_named, useNA = "ifany")
```

# Per-phase ISFI

`compute_isfi()` with `phase_col` returns one row per
(individual, phase) combination — exactly the breakdown needed for
biological interpretation:

```{r}
isfi_by_phase <- compute_isfi(phs,
                               time_col       = "time",
                               individual_col = "individual",
                               cluster_col    = "cluster_id",
                               phase_col      = "phase_named")
print(isfi_by_phase)
```

# Fidelity heatmap

A compact way to read the per-phase fidelities is a heatmap: one
column per individual, one row per phase ordered **Breeding (top) →
Stopover → Wintering (bottom)**, so the breeding-ground fidelity sits
at the top. The demo carries a **single** individual, so the map is
one column wide; on the full nine-individual dataset each bird adds a
column, making between-individual differences immediately legible.

```{r}
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Inter-annual site fidelity by life-history phase. Warm = high fidelity, cool = low."
phase_order <- c("wintering", "stopover", "breeding")   # bottom -> top
hm <- isfi_by_phase
hm$phase      <- factor(hm$phase, levels = phase_order)
hm$individual <- factor(hm$individual)

ggplot(hm, aes(individual, phase, fill = isfi)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(is.na(isfi), "", sprintf("%.2f", isfi))),
            size = 3.4, colour = "grey15") +
  scale_fill_distiller(name = "fidelity", palette = "RdYlBu", direction = -1,
                       limits = c(0, 1), breaks = c(0, 0.5, 1),
                       labels = c("0.0 (Low)", "0.5 (Mid)", "1.0 (High)"),
                       na.value = "grey88") +
  labs(x = "individual", y = NULL,
       title = "Site fidelity by phase") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Interpretation

Black-tailed Godwits are long-distance migrants with strong site
fidelity. The expected pattern across phases:

- **Breeding** — highest ISFI; godwits return to the same nesting
  territory year after year.
- **Wintering** — usually high; the species is faithful to specific
  coastal mudflats and lagoons.
- **Stopover** — most variable; choice of refuelling sites depends
  on weather, food availability, and timing.

The numerical breakdown above quantifies these patterns for this
specific individual.

# INFI: nest-level fidelity

The same function applied to the output of `detect_nests()` yields
the Individual Nest Fidelity Index:

```{r eval = FALSE}
nests <- detect_nests(breeding_pts,
                      time_col       = "time",
                      individual_col = "individual")$nests
infi  <- compute_isfi(nests,
                      time_col       = "start_time",
                      individual_col = "individual",
                      cluster_col    = "nest_id")
```

# See also

- `vignette("v04-mpi-phases")` — phase classification (upstream)
- `vignette("v07-st-dbscan-nesting")` — produces the input for INFI

```{r}
sessionInfo()
```
