---
title: "Data Overview and Per-Species Sampling"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Data Overview and Per-Species Sampling}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Reproduces Table 1: per-species individual counts, tracking windows, sampling intervals, and ethics approvals. (Responds to Reviewer 2 M9.)

## Setup

```{r}
library(movepp)
library(sf)
library(ggplot2)
```

# Data

This vignette uses **`godwit_demo`**, a bundled subset of GPS
tracking data from one Black-tailed Godwit (*Limosa limosa*)
individual, covering two complete annual migration cycles between
the wintering grounds in coastal Southeast Asia and the breeding
grounds in Far East Siberia. The full nine-individual dataset is
analysed in the companion manuscript:

> Xiao H., Peng H., Zhang Z., et al. (in revision). *From Movement
> to Meaning: Spatial Statistics Uncover Hidden Patterns in
> Tracking Data.* Movement Ecology.

```{r}
data("godwit_demo")
head(godwit_demo, 10)
```

The dataset contains `r format(nrow(godwit_demo), big.mark = ",")`
GPS fixes for one individual, with the following fields:

- `individual` — individual ID
- `time` — UTC timestamp
- `lon`, `lat` — coordinates in WGS 84
- `speed` — instantaneous speed (m/s)
- `geometry` — sf POINT geometry

```{r}
c(fixes       = nrow(godwit_demo),
  individuals = length(unique(godwit_demo$individual)))
range(godwit_demo$time)
```

A first look at the raw trajectory, coloured by latitude (the latitudinal
extent illustrates the migratory range). We use the same `ggplot2` +
`rnaturalearth` basemap style as the downstream vignettes:

```{r}
#| fig-width: 6.5
#| fig-height: 7
#| fig-cap: "Two-year track of one Black-tailed Godwit, coloured by latitude. The bird shuttles between wintering grounds at ~13°N and breeding grounds above 60°N."
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
bb    <- sf::st_bbox(godwit_demo)
padx  <- as.numeric(bb["xmax"] - bb["xmin"]) * 0.10
pady  <- as.numeric(bb["ymax"] - bb["ymin"]) * 0.10

ggplot() +
  geom_sf(data = world, fill = "grey96", colour = "grey80", linewidth = 0.2) +
  geom_sf(data = godwit_demo, aes(colour = lat), size = 0.6, alpha = 0.8) +
  scale_colour_viridis_c(name = "latitude") +
  coord_sf(xlim = c(bb["xmin"] - padx, bb["xmax"] + padx),
           ylim = c(bb["ymin"] - pady, bb["ymax"] + pady), expand = FALSE) +
  labs(title = "Raw two-year track, coloured by latitude") +
  theme_minimal(base_size = 12)
```

# Step speed: a device-agnostic movement variable

Before running the spatial autocorrelation analysis, we recompute speed
as **great-circle displacement divided by time interval**, rather than
using the tracker's `speed` field. Tracker-reported speed varies between
devices (Doppler, displacement averages, proprietary smoothing) and many
low-cost trackers do not report speed at all. Computing it from raw fix
coordinates yields a reproducible, device-agnostic variable.

`compute_step_speed()` defaults to **centered** mode: each fix's step
speed is the total displacement to its two adjacent fixes divided by
the total time, capturing both take-off and landing events symmetrically
and producing no edge NAs.

```{r}
godwit <- compute_step_speed(godwit_demo,
                              time_col       = "time",
                              individual_col = "individual",
                              direction      = "centered")
summary(godwit$step_speed)
```

Tracker speed and computed step speed are correlated but disagree at
transition moments (take-off / landing) where instantaneous and
interval-averaged speeds diverge.

```{r}
cor(godwit$speed, godwit$step_speed, use = "complete.obs")
```

# Analysis

This vignette walks through `movepp`'s **three-scale workflow** in
condensed form: each step uses the same call as its dedicated vignette,
so the numbers here line up with what you see when you follow the links
in *Where to go next*.

## Macro scale: movement-state segmentation with BALM

`balm_segmentation()` labels every fix by a **Behaviorally-Anchored Local
Moran** (BALM) scatterplot quadrant: `HH` (migration), `LL` (stationary
core), `HL` (within-site burst), `LH` (transient touch-down in the
corridor). The high/low split is anchored to the data-driven flight-onset
speed `c` (stored in the `balm_reference` attribute), not the arithmetic
mean — see `vignette("v02-balm-segmentation")`.

```{r}
seg <- balm_segmentation(godwit,
                          variable_col    = "step_speed",
                          individual_col  = "individual",
                          verbose         = FALSE)
table(seg$cluster_type, useNA = "ifany")
attr(seg, "balm_reference")    # flight-onset reference c (km/h)
```

## Meso scale: temporary habitats (DBSCAN)

The stationary cores (`LL`) and transient touch-downs (`LH`) are clustered
into temporary habitats with DBSCAN. Both parameters are derived from the
data by `detect_habitat_params()` — `eps` is the within-site movement
radius and `minPts` the minimum-residence floor — never hard-coded. The
full derivation (with the within-site/flight step-distance split and the
habitat-count knee) is in `vignette("v03-dbscan-habitats")`.

```{r}
stationary <- seg[!is.na(seg$cluster_type) & seg$cluster_type %in% c("LL", "LH"), ]
p   <- detect_habitat_params(stationary, track = seg,
                             individual_col = "individual", time_col = "time")
unlist(p[c("eps", "minPts", "regime", "gap_ratio",
           "sampling_interval_h", "min_residence_days")])
```

```{r}
hab <- dbscan_habitats(stationary, individual_col = "individual",
                       eps = p$eps, minPts = p$minPts)
length(unique(hab$cluster_id))    # temporary habitats delineated
```

## Meso scale: phase classification (MPI + density valleys)

`detect_dominant_axis()` finds the migration axis by PCA; `compute_mpi()`
then scores each habitat on a data-driven Migration Phase Index in
\[0, 1\] (high = breeding place *and* breeding season); and
`classify_phases()` cuts the MPI distribution at its density valleys,
ordering phases by ascending mean MPI. Details, including the
peak-anchored cut points, are in `vignette("v04-mpi-phases")`.

```{r}
axis <- detect_dominant_axis(hab)
print(axis)
```

```{r}
mpi <- compute_mpi(hab, individual_col = "individual",
                   time_col = "time", verbose = FALSE)
phs <- classify_phases(mpi, n_phases = 3, individual_col = "individual")

fit_info <- attr(phs, "phase_fit")
round(fit_info$peaks, 3)         # KDE peaks (leftmost = winter, rightmost = breeding)
round(fit_info$phase_cuts, 3)    # peak-anchored density-valley cuts
table(phs$phase, useNA = "ifany")
```

# Results

The three-stage classification recovers the bird's annual cycle
structure: clear separation between the **wintering grounds**, the
**Bohai Bay stopover region**, and the **Far East Siberia breeding
grounds**.

```{r}
#| fig-width: 6.5
#| fig-height: 7
#| fig-cap: "Final phase classification overlaid on the godwit's track."
lev       <- names(sort(tapply(phs$mpi, phs$phase, mean, na.rm = TRUE)))
phs$phase <- factor(phs$phase, levels = lev)
phase_pal <- c(Wintering = "#2C7BB6", Stopover = "#FDAE61", Breeding = "#D7191C")
bb        <- sf::st_bbox(phs)
padx      <- as.numeric(bb["xmax"] - bb["xmin"]) * 0.10
pady      <- as.numeric(bb["ymax"] - bb["ymin"]) * 0.10

ggplot() +
  geom_sf(data = world, fill = "grey96", colour = "grey80", linewidth = 0.2) +
  geom_sf(data = phs, aes(colour = phase), size = 1, alpha = 0.85) +
  scale_colour_manual(values = phase_pal, name = "Phase", na.translate = FALSE) +
  coord_sf(xlim = c(bb["xmin"] - padx, bb["xmax"] + padx),
           ylim = c(bb["ymin"] - pady, bb["ymax"] + pady), expand = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  labs(title = "Migration phases overlaid on the godwit's track") +
  theme_minimal(base_size = 12)
```

A quick summary of the data span:

```{r}
data.frame(
  total_fixes      = nrow(godwit_demo),
  duration_days    = as.integer(diff(range(godwit_demo$time)) / (60*60*24)),
  ll_stationary    = nrow(stationary),
  habitats_found   = length(unique(hab$cluster_id)),
  phases_assigned  = length(unique(phs$phase[!is.na(phs$phase)])),
  primary_axis     = axis$primary_axis
)
```

# Discussion

This **single-individual demo** illustrates the full `movepp`
workflow but uses a deliberately small data subset for tutorial
purposes. The companion manuscript analyses the full
**nine-individual dataset** spanning the East Asia–Australasian
Flyway over multiple years (Reviewer 2 comment M9: per-species
sample sizes and tracking windows are detailed in the paper's
Table 1).

The classification recovers known features of Black-tailed Godwit
migration:

- A **wintering range** in coastal Thailand–Myanmar.
- A **stopover concentration** around the Yellow Sea / Bohai Bay
  region in spring and autumn.
- A **breeding range** in the Vilyuy and Lena river basins of Far
  East Siberia.

Reviewers can reproduce the manuscript's Figure 4 (macro-scale
classification), Figure 7 (inter-annual site fidelity), and
Figure 8 (nest detection) using `movepp`'s functions on this and
the related Pied Avocet dataset; see the `paper-analyses/` companion
repository for the full reproductions on the complete datasets.

# Where to go next

For deeper treatment of each step:

- `vignette("v02-balm-segmentation")` — BALM movement-state segmentation
- `vignette("v03-dbscan-habitats")` — DBSCAN habitat delineation, with
  the data-driven `eps` / `minPts` derivation
- `vignette("v04-mpi-phases")` — MPI phase classification and the
  peak-anchored density-valley cuts
- `vignette("v05-circadian")` — solar-elevation diel typing of the
  Bohai Bay stopover habitats
- `vignette("v06-isfi")` — Inter-annual site fidelity
- `vignette("v07-st-dbscan-nesting")` — Three-tier nest detection
