---
title: "MPI Phase Classification"
format:
  html:
    embed-resources: true
    toc: true
vignette: >
  %\VignetteIndexEntry{MPI Phase Classification}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

# Overview

The meso-scale phase chain assigns each temporary habitat to a life-history
phase (wintering / stopover / breeding):

1. **`compute_mpi()`** -- a **data-driven** Migration Phase Index (MPI) in
   \[0, 1\]: nothing is hard-coded to latitude or to a fixed calendar.
2. **`classify_phases()`** -- finds the **density valleys** of the MPI
   distribution and **cuts** the MPI range at them, labelling each point by
   the band it falls in, then votes per habitat; phases are ordered by
   ascending mean MPI.
3. **`annotate_phases()`** -- an optional interactive gadget to manually
   refine the automatic result (seeded from it), per habitat.

`detect_dominant_axis()` is available as a stand-alone axis diagnostic.

# Setup

```{r}
library(movepp); library(sf); library(ggplot2)
data("godwit_demo")
godwit <- compute_step_speed(godwit_demo, time_col = "time",
                             individual_col = "individual", direction = "centered")
seg <- balm_segmentation(godwit, variable_col = "step_speed",
                        individual_col = "individual", verbose = FALSE)
stationary <- seg[!is.na(seg$cluster_type) & seg$cluster_type %in% c("LL", "LH"), ]
p <- detect_habitat_params(stationary, track = seg,
                        individual_col = "individual", time_col = "time")
hab <- dbscan_habitats(stationary, individual_col = "individual",
                       eps = p$eps, minPts = p$minPts)
```

# The Migration Phase Index

MPI is the geometric mean of a **spatial** term and a **temporal** term, so a
high value requires being at the breeding *place* AND in the breeding *season*
-- an estimate of the chance of breeding. Everything is derived from the data:

* **Axis** -- PCA of the (lon, lat) coordinates; PC1 is the dominant migration
  direction (latitudinal, longitudinal or diagonal). If PC1 explains < 85% of
  variance, PC2 is added (three-way geometric mean).
* **Polarity & temporal anchors (population)** -- for each well-travelled
  individual the lowest / highest 10% (`q`) of axis positions form two extreme
  fix-sets. Their pooled occupancy **dates** do two jobs: they decide which end
  is breeding (the extreme whose dates are more temporally *concentrated* -- the
  short, synchronised breeding window; a poleward cross-check breaks ties), and
  they set the population wintering / breeding day-of-year anchors. Pooled
  across individuals -- "time = population".
* **Spatial term (per individual)** -- each individual's axis position is
  rescaled to \[0, 1\] between its **near-maximal** extremes (the 2nd / 98th
  percentiles, `anchor_q`), **not** the 10% tails used above. The breeding
  extreme is a *spatial* limit that can be badly under-represented in *time*: an
  individual that skips breeding in some years spends most of its northern time
  at a lower, non-breeding settlement, which would pull a 10th/90th-percentile
  anchor down and clamp the true breeding extreme to 1 -- making a non-breeding
  stop indistinguishable from real breeding. Near-maximal anchors reserve
  `MPI = 1` for the genuine extreme. "Space = individual".
* **Temporal term** -- a triangular tent: 0 at the population wintering date,
  rising linearly to 1 at the breeding date, then falling back to 0, with
  possibly unequal spring / autumn arms. Spring and autumn map to the same
  value, so MPI is a direction-agnostic breeding-*phase* index.

`winter_doy` / `breed_doy` can be supplied to override the auto-detected
anchors (e.g. tropical or photoperiod-decoupled breeders).

```{r}
mpi <- compute_mpi(hab, individual_col = "individual", time_col = "time")
summary(mpi$mpi)
c(winter_doy = attr(mpi, "mpi_winter_doy"),
  breed_doy  = attr(mpi, "mpi_breed_doy"),
  pc1_var    = round(attr(mpi, "mpi_pc1_var"), 3))
```

# Phase classification

`classify_phases()` fits **no distributional model** and uses **no prominence
threshold**. MPI is built so wintering piles up near 0 and breeding near 1, so
on the pooled MPI density the **leftmost peak is the wintering mode and the
rightmost peak is the breeding mode** -- a structural fact of the index. The two
phase boundaries are anchored to those extreme peaks: the **lower cut** is the
first density valley to the right of the leftmost peak, and the **upper cut** is
the last density valley to the left of the rightmost peak. Everything between is
the stopover band, so **any number of staging sites -- however deep their
internal valleys -- are absorbed into stopover and can never be mistaken for a
phase boundary**. Each point is then labelled **deterministically** by the band
its MPI falls in, and each habitat takes the **majority** phase of its points.
The resolved phase count shrinks below the request only when the data
structurally lack the modes (two peaks -> two phases; one mode -> one) -- an
objective property of the distribution, not a tuned cutoff.

```{r phase-3}
phs <- classify_phases(mpi, n_phases = 3, individual_col = "individual")

lev <- names(sort(tapply(phs$mpi, phs$phase, mean, na.rm = TRUE)))
phs$phase <- factor(phs$phase, levels = lev)
phase_pal <- c(Wintering = "#2C7BB6", Stopover = "#FDAE61", Breeding = "#D7191C")
fit_info <- attr(phs, "phase_fit")
round(fit_info$peaks, 3)             # KDE peaks (leftmost = winter, rightmost = breeding)
round(fit_info$phase_cuts, 3)        # peak-anchored cut points
fit_info$phase_sizes                 # points per phase
round(fit_info$phase_mpi_mean, 3)    # mean MPI per phase
table(phs$phase)
```

# MPI distribution by phase

The histogram shows the "spike-plateau-spike" MPI distribution coloured by the
assigned phase, with the **peak-anchored cuts** (dashed) marked. The lower cut
sits just right of the wintering peak, the upper cut just left of the breeding
peak; because assignment is a hard cut, each colour boundary falls exactly on a
dashed line.

```{r mpi-histogram, fig.width=7.5, fig.height=4.5}
cuts <- attr(phs, "phase_fit")$phase_cuts
cuts <- cuts[is.finite(cuts)]
hdf  <- data.frame(mpi = phs$mpi, phase = phs$phase)
hdf  <- hdf[!is.na(hdf$mpi), ]
ggplot(hdf, aes(x = mpi, fill = phase)) +
  geom_histogram(bins = 40, colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = cuts, linetype = "dashed", colour = "grey30") +
  scale_fill_manual(values = phase_pal, name = "Phase") +
  labs(x = "Migration Phase Index (MPI)", y = "Count",
       title = "MPI distribution by phase (dashed = density-valley cuts)") +
  theme_minimal(base_size = 12)
```

# Spatial layout by phase

```{r phase-spatial, fig.width=6.5, fig.height=7}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
bb <- sf::st_bbox(phs)
padx <- as.numeric(bb["xmax"] - bb["xmin"]) * 0.10
pady <- as.numeric(bb["ymax"] - bb["ymin"]) * 0.10
ggplot() +
  geom_sf(data = world, fill = "grey96", colour = "grey75", linewidth = 0.2) +
  geom_sf(data = phs, aes(colour = phase), size = 2, alpha = 0.85) +
  scale_colour_manual(values = phase_pal, name = "Phase") +
  coord_sf(xlim = c(bb["xmin"] - padx, bb["xmax"] + padx),
           ylim = c(bb["ymin"] - pady, bb["ymax"] + pady), expand = FALSE) +
  labs(title = "Spatial distribution of migration phases") +
  theme_minimal(base_size = 12)
```

# Manual refinement (optional, interactive)

`annotate_phases()` opens a Shiny gadget **seeded with the phase classification**
(`phase_col = "phase"`): a habitat-occupancy *behaviour barcode* (time x habitat
rank) linked to a clean coastline map. Box-select a habitat -- a rank-row over a
time span -- on either panel and reassign its phase; the corrections come back
in a `phase_manual` column. Because the barcode is organised by habitat, edits
are made **per habitat**, matching the cluster-level classification, so only the few
mis-classified habitats need fixing rather than annotating from scratch.

The gadget needs a per-individual habitat rank (south -> north) for the barcode
y-axis:

```{r}
co <- sf::st_coordinates(phs); lat <- co[, 2]
phs$cluster_rank <- NA_integer_
for (id in unique(phs$individual)) {
  s    <- phs$individual == id
  mlat <- tapply(lat[s], phs$cluster_id[s], mean)
  rk   <- rank(mlat, ties.method = "first")
  phs$cluster_rank[s] <- as.integer(rk[as.character(phs$cluster_id[s])])
}
```

The call below launches the interactive gadget, so it is not run when building
this vignette:

```{r}
#| eval: false
phs <- annotate_phases(phs, time_col = "time", rank_col = "cluster_rank",
                       individual_col = "individual", phase_col = "phase")
# refined labels are in phs$phase_manual
table(phs$phase_manual)
```

# See also

- `vignette("v03-dbscan-habitats")` -- upstream habitat delineation
- `vignette("v06-isfi")` -- using phase labels for stratified site fidelity
```