---
title: "Nest Detection with Stratified ST-DBSCAN"
format:
  html:
    embed-resources: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Nest Detection with Stratified ST-DBSCAN}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

# Overview

`detect_nests()` implements the **three-tier sigma-stratified
ST-DBSCAN** algorithm of the accompanying manuscript (Eq. 10). It
exploits the fact that GPS fixes during nesting are approximately
bivariate-normally distributed around the true nest location:

- ~95% of fixes within 2σ of the centre
- ~68% within 1σ
- ~38% within 0.5σ

A point is classified as a nest fix only if it survives the density
threshold at all three sigma tiers simultaneously, yielding markedly
fewer false positives than single-tier DBSCAN.

# Recover the breeding subset from the upstream pipeline

Rather than hard-coding latitude or month thresholds, we reuse the
phase classification from vignettes 01-04: the breeding-phase
habitats are the ones with the highest mean MPI score in the phase classification
voting output.

```{r}
library(movepp)
library(sf)
library(ggplot2)
data("godwit_demo")

godwit <- compute_step_speed(godwit_demo,
                              time_col       = "time",
                              individual_col = "individual",
                              direction      = "centered")

seg <- balm_segmentation(godwit,
                          variable_col    = "step_speed",
                          individual_col  = "individual",
                          verbose         = FALSE)

stationary   <- seg[!is.na(seg$cluster_type) & seg$cluster_type %in% c("LL", "LH"), ]
p    <- detect_habitat_params(stationary, track = seg,
                              individual_col = "individual", time_col = "time")
hab  <- dbscan_habitats(stationary,
                        individual_col = "individual",
                        eps = p$eps, minPts = p$minPts)
mpi  <- compute_mpi(hab, individual_col = "individual",
                    time_col = "time", verbose = FALSE)
phs  <- classify_phases(mpi, n_phases = 3, individual_col = "individual")

# Map ascending-MPI phase labels to biological names
ph_df    <- sf::st_drop_geometry(phs)
ph_order <- names(sort(tapply(ph_df$mpi, ph_df$phase, mean, na.rm = TRUE)))
phase_map <- setNames(c("wintering", "stopover", "breeding"), ph_order)
phs$phase_named <- factor(phase_map[as.character(phs$phase)],
                          levels = c("wintering", "stopover", "breeding"))

breeding <- phs[!is.na(phs$phase_named) & phs$phase_named == "breeding", ]

cat("Breeding-phase fixes:", nrow(breeding), "\n")
cat("Date range:", format(range(breeding$time)), "\n")
```

# Project to a metric CRS

`detect_nests()` requires coordinates in metres so that `search_dist`
has a physical interpretation. UTM 52N covers the Vilyuy / Lena
breeding range:

```{r}
breeding_utm <- sf::st_transform(breeding, 32652)
```

# Run detect_nests

With a 3-hour sampling interval, ~30 fixes cover roughly 4 days of
continuous nest attendance — a reasonable lower bound. `search_dist`
is relaxed to 50 m to absorb GPS positional noise on individual
fixes.

```{r}
res <- detect_nests(breeding_utm,
                    time_col       = "time",
                    individual_col = "individual",
                    min_pts        = 30L,
                    search_dist    = 50,
                    time_window    = "3 weeks",
                    verbose        = FALSE)

res$nests
```

```{r}
cat("Fixes assigned to a nest:", sum(!is.na(res$points$nest_id)), "\n")
cat("Foraging-area fixes:     ", sum( is.na(res$points$nest_id)), "\n")
```

# Map: detected nests on the breeding grounds

We transform the nest centroids and the classified fixes back to
WGS 84 and map them over the Far East Siberian breeding range
(60–70 °N, 90–140 °E). Nest fixes (red) form tight cores within the
looser cloud of foraging fixes (grey); each nest centroid is labelled
with its `nest_id`.

```{r}
#| fig-width: 7.5
#| fig-height: 6
#| fig-cap: "ST-DBSCAN nest detections on the Siberian breeding grounds. Red points are nest fixes, grey are foraging fixes; labels mark each nest centroid."
world    <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
pts_ll   <- sf::st_transform(res$points, 4326)
pts_ll$is_nest <- !is.na(pts_ll$nest_id)
nests_ll <- sf::st_transform(res$nests, 4326)
nc       <- sf::st_coordinates(nests_ll)
nests_ll$lon <- nc[, 1]; nests_ll$lat <- nc[, 2]

ggplot() +
  geom_sf(data = world, fill = "grey96", colour = "grey80", linewidth = 0.2) +
  geom_sf(data = pts_ll, aes(colour = is_nest), size = 0.7, alpha = 0.6) +
  geom_sf(data = nests_ll, shape = 21, fill = "#D7191C",
          colour = "white", size = 3, stroke = 0.6) +
  geom_text(data = nests_ll, aes(lon, lat, label = nest_id),
            nudge_y = 0.6, nudge_x = 0.6, size = 3.4,
            fontface = "bold", colour = "grey15") +
  scale_colour_manual(values = c(`TRUE` = "#D7191C", `FALSE` = "grey60"),
                      labels = c(`TRUE` = "nest fix", `FALSE` = "foraging fix"),
                      name = NULL) +
  coord_sf(xlim = c(90, 140), ylim = c(60, 70), expand = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "Detected nests on the Far East Siberian breeding grounds",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12)
```

# Why three tiers?

Default `sigma_ratios = c(1.0, 0.68, 0.38)` (paired with internal
distance multipliers `c(1.0, 0.5, 0.25)`) follow the bivariate-normal
geometry of GPS fixes around a true nest. The stratified test —
density requirements satisfied at ALL three radii — sharply
discriminates true nest fixes from loose foraging clusters that
might satisfy a single-tier test but lack the tight inner core.

# Synthetic-data illustration

For a clean algorithm demonstration with known ground truth, simulate
a nest cluster surrounded by foraging fixes:

```{r eval = FALSE}
set.seed(42)
n_nest <- 200
n_off  <- 100
pts_syn <- sf::st_as_sf(data.frame(
  lon  = c(rnorm(n_nest, 0, 3), rnorm(n_off, 0, 150)),
  lat  = c(rnorm(n_nest, 0, 3), rnorm(n_off, 0, 150)),
  time = seq.POSIXt(as.POSIXct("2023-05-15"),
                    by = "1 hour", length.out = 300),
  individual = "synthetic_bird"
), coords = c("lon", "lat"), crs = 32633)

detect_nests(pts_syn, "time", "individual",
             min_pts = 40L, search_dist = 30,
             time_window = "3 weeks")$nests
```

With hourly sampling and a tight 3-m cluster, the algorithm recovers
the nest centre to sub-metre precision. On real 3-hour-interval data
the parameters are relaxed (50 m, 30 points) to accommodate GPS
noise and the coarser temporal grid.

# Cross-validate with `nestR`

```{r eval = FALSE}
remotes::install_github("picardis/nestR")
library(nestR)
nestR_result <- find_nests(gps_data = your_data, buffer = 10)
```

A side-by-side accuracy comparison on labelled data is provided in
the `paper-analyses/` companion repository.

# See also

- `vignette("v04-mpi-phases")` — produces the phase labels (upstream)
- `vignette("v06-isfi")` — uses the `nests` output to compute INFI

```{r}
sessionInfo()
```
